---
title: "Get a look at that hazard!"
author: "Ben Ogorek"
header-includes:
   - \usepackage{bbm}
date: "May 19, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Distributions vs Hazards 

Distributions are hard things to love. They admit uncertainty.
 They entail extra work beyond the best guess from an expectation or 
machine learning model. They involve often dubious assumptions and 
they can look rediculous in hindsight (TODO: add link).

Work in "slight of hand" 

That's why I've been surprized that modeling the hazard,
 itself nothing more than an alternative representation
of a distribution of lifetimes, is more appealing to dabblers in
statistics. Maybe there's something fundamental about the uncertainty in when
something will end, but I've met more than one analyst who really wanted to
work with hazards. 


I always push newcomers to discrete time-to-event analysis (and especially
 [this excellent paper][https://statisticalhorizons.com/wp-content/uploads/Allison.SM82.pdf],
by Paul Allison), but the dabblers push back and want the real thing.
The celebrated proportional hazards model,
(Cox 1972)[1][1]
$$
\lambda_0(t) \exp{\mathbf{x}_i * \beta}
$$
...
where $\lambda_0(t)$ is the continuous "baseline" hazard function. Since
we've subscribed to the proportional hazards assumption (and barring any
time-varying covariates for now), we just need to look at $\lambda_0(t)$
to get an understanding of the instantaneous risk through time. Does it have
a bathtub shape? Is it always increasing? Or is it mostly flat?

These are interesting and useful questions that shed light on the failure
mechanism. So why then, does nearly every survival analysis tutorial on
the web focus on estimating $H(t) = \int_0^t \lambda_0(t) \dt$, the "cumulative
hazard", without ever showing $lambda_0(t)$? 

Let's take it from a bayesian non-parametrics guy (Hjort 1990)[2][2]) who
shouldn't be afraid to do anything:
"Now \[the hazard\] is as difficult to estimate with good precision as is the
density $F^{\prime}(t)$, so we prefer working with with the *cumulative 
hazard*...The cumulative hazard can be defined even when $F$ has no density."


In this article, we're going to *see* the hazard. Perhaps you're saying to
yourself, if you've got the integral of the quantity you want, you're 90%
of the way there - just use an approximate differentiation method. And we
will, though we'll have to do some smoothing first. We'll also simulate
from a baseline hazard that you probably never thought you'd come across.

## Simulation

Beyond R, there's likely nothing to be installed packages for this tutorial,
as we'll be depending on `survival` and `splines` packages, a *recommended*
and *base* package, respectively).
```{r}
library(survival)
library(splines)
```

First, let's choose our sample size and set up an artificial data set with
two covariates, $x_1$ and $x_2$.

```{r}
N <- 1000
sim_data <- data.frame(x1 = rnorm(N), x2 = rnorm(N))
```

Though we're working with continuous time methods, for the purposes of
simulation we must discretize time:

```{r}
t_max_for_sim <- 25 # Go out far enough to exceed last event (censor or death)
delta <- .01

t_seq <- seq(0, t_max_for_sim, delta)
```

Now here is the baseline hazard we're going to be simulating from. It's 
a sinosoid with a bias term and a period of 12. Not exactly a "bathtub":

```{r}
baseline_hazard <- .1 * (1.1 + sin(2 * pi * t_seq / 12))
plot(baseline_hazard ~ t_seq)
```

```{r}
simulate_lifetime <- function(x1, x2) {
  i <- 0
  alive <- TRUE
  censored <- FALSE
  t_max_observable <- rgamma(1, 25, 2)
  while(alive & !censored) {
    i <- i + 1
    pr_die_in_interval <- delta * baseline_hazard[i] * exp(.1 * x1 - .1 * x2)
    alive <- !(runif(1) <= pr_die_in_interval)
    censored <- t_seq[i] >= t_max_observable
  }
  return(c(t_seq[i], censored))
}

dat <- sapply(1:N, FUN = function(i) with(sim_data[i, ],
                                          simulate_lifetime(x1, x2)))
sim_data <- as.data.frame(cbind(sim_data, t(dat)))
names(sim_data) <- c("x1", "x2", "t", "censored")

# Start analysis with CoxPh

my_model <- coxph(Surv(time = t, event = !censored, type = 'right') ~ x1 + x2,
                  data = sim_data)
summary(my_model)
baseline_surv <- survfit(my_model)

surv_df <- data.frame(S = baseline_surv$surv,
                      H = baseline_surv$cumhaz,
                      t = baseline_surv$time)


surv_spline <- lm(H ~ ns(t, df = 12), data = surv_df)
plot(surv_df$H ~ surv_df$t)
lines(surv_spline$fitted ~ surv_df$t, col = "red")

grid_df <- data.frame(t = t_seq)
grid_df$H <- predict(surv_spline, newdata = grid_df)

plot(diff(grid_df$H) / delta ~ t_seq[-1], xlim = c(0, 18))
lines(baseline_hazard[-1] ~ t_seq[-1], col = "red")
```



## References

[1]: Cox, D. R. 1972. "Regression Models and Life Tables (with Discussion)." Journal of the Royal Statistical
Society, Series B 34:187-220.

[2]: Hjort, N. (1990). Nonparametric Bayes estimators based on beta processes in models for life history data. Annals of Statistics 18, 1259–1294.
